---
title: "In-depth Analysis of School Nutrition Programs and Dietary Intake"
author: "Prajin Dhumal"
date: "04/07/2024"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

The primary focus of my report is to examine the association between school nutrition agendas and student dietary patterns. Comprehending this connection is essential as it can help inform educational policy and program design to improve student health outcomes.

Current policy shifts have produced a need for strict analysis to define the effectiveness of these programs. To manage this gap, my analysis leverages data from school nutrition records and national health surveys.

## Research Question

My analysis looks to answer the following question: How do school nutrition programs impact students' dietary behaviors and nutritional status?

## Importance of the Study

Investigating this topic is very important for multiple stakeholders, including policymakers, school administrators, and health professionals, as it provides evidence-based insights that can drive decisions in educational and health sections.


## The Two Data Sources
```{r}
school_nutrition <- read.csv("C:/Users/praji/School_Nutrition_Program_Year_20212022.csv")
head(school_nutrition)

nhanes_diet <- read.csv("C:/Users/praji/Dietary_Intake_Estimates.csv")
head(nhanes_diet)
```


## Data Inspection
```{r}
# School Nutrition Programs Data Inspection
str(school_nutrition)
glimpse(school_nutrition)

# NHANES Dietary Intake Estimates Inspection
str(nhanes_diet)
glimpse(nhanes_diet)
```

## Initial Data Cleaning
```{r create-school-nutrition-clean}
# Creating the school_nutrition_clean dataframe from school_nutrition
school_nutrition_clean <- school_nutrition %>%
  filter(!is.na(TotalMealsSnacks)) %>%  # Ensure there are no NA values in the TotalMealsSnacks column
  mutate(PerCapitaMeals = TotalMealsSnacks / EnrollmentQty)  # Calculate meals served per capita
```

## Data Wrangling

### Enhancing School Nutrition Data with Regular Expressions and User-Defined Functions

```{r enhance-school-nutrition}
# Example use of a regular expression to extract data from a string column
school_nutrition_clean <- school_nutrition_clean %>%
  mutate(FirstWordCEName = str_extract(CEName, "^\\w+"))

# User-defined function to calculate total meals served
calculate_total_meals <- function(breakfast, lunch, snacks) {
  breakfast + lunch + snacks
}

# Use the user-defined function with mutate and rowwise for row-wise operations
school_nutrition_clean <- rowwise(school_nutrition_clean) %>%
  mutate(TotalMealsServed = calculate_total_meals(BreakfastServedFree, LunchServedFree, SnacksServedFree))

# Now ungroup the data after rowwise operation
school_nutrition_clean <- ungroup(school_nutrition_clean)
```


### Demonstrating Control Flow with Loops

```{r}
# Normalize ADP columns by the maximum value for comparison
normalized_columns <- c("BreakfastADP", "LunchADP", "SnackADP")
for (col in normalized_columns) {
  max_value <- max(school_nutrition_clean[[col]], na.rm = TRUE)
  school_nutrition_clean[[col]] <- school_nutrition_clean[[col]] / max_value
}
```


### Preparing NHANES Dietary Data

```{r}
# Filter to include only data from relevant years and clean specific nutrient columns
nhanes_clean <- nhanes_diet %>%
  filter(`Survey.Years` == "2017-2018")

# Pivot data from wide to long format to better handle nutrient variables
nhanes_long <- pivot_longer(nhanes_clean,
                            cols = c(`Mean`, `Standard.Error`, `Lower.95..CI.Limit`, `Upper.95..CI.Limit`),
                            names_to = "Statistic_Type",
                            values_to = "Value")

# Summarize the long data to get average values per nutrient statistic
nhanes_summary <- nhanes_long %>%
  group_by(Survey.Years, Statistic_Type) %>%
  summarise(AverageValue = mean(Value, na.rm = TRUE), .groups = 'drop')

knitr::kable(nhanes_summary, caption = "Average Nutrient Statistics by Year and Type", format = "html")
```


## Data Visualization

### Relationship Between Total Meals Served, Reimbursement, and Enrollment Over Years

```{r visualization-complex, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(school_nutrition_clean, aes(x = TotalMealsServed, y = TotalReimbursement)) + 
  geom_point(aes(color = as.factor(ProgramYear), size = EnrollmentQty), alpha = 0.7) +  # Points with color and size representing different variables
  geom_smooth(aes(color = as.factor(ProgramYear)), method = "lm", se = FALSE) +  # Regression line per ProgramYear
  scale_color_brewer(palette = "Set1") +  # Color palette for the ProgramYear
  scale_size(range = c(1, 8)) +  # Size range for EnrollmentQty
  labs(title = "Total Meals Served vs. Reimbursement by Year",
       subtitle = "Size of points represents enrollment quantity",
       x = "Total Meals Served",
       y = "Total Reimbursement",
       color = "Program Year",
       size = "Enrollment Qty") +
  theme_minimal() +
  theme(legend.position = "bottom")  # Adjust legend position
```


### Visualization for NHANES Dietary Data

```{r}
ggplot(nhanes_summary, aes(x = Statistic_Type, y = AverageValue)) + 
  geom_bar(aes(fill = as.factor(Survey.Years)), stat = "identity", position = "dodge") +
  labs(title = "Average Nutrient Statistics by Year and Type",
       x = "Nutrient Statistic Type",
       y = "Average Value",
       fill = "Survey Year") +
  theme_minimal()
```

This analysis uses two primary datasets: School Nutrition Program Data for the 2021-2022 academic year, and NHANES Dietary Intake Estimates. The datasets are available in the readme.md file





